Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-08-09 manuscript.qmd

1. We assume that both ipsilateral and contralateral spread is described through the same graph shown in {#fig-small-graph}.
2. We assume the spread *among* the LNLs to be same on both sides. It is reasonable to assume the lymphatic system is symmetric. Thus, the spread rates from one LNL to the other should be symmetric, too. Formally, this means \
   $$
   \begin{aligned}
   b_v^\text{c} &\neq b_v^\text{i} \\
   t_{rv}^\text{c} &= t_{rv}^\text{i}
   \end{aligned}
   $$ {#eq-symmetries}
   for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
3. The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side. This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to the same position, but on the contralateral side. In the beginning we would have $b_v^\text{c} < b_v^\text{i}$, while in the end, the situation is reversed. If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes. We encode this in a *mixing parameter* $\alpha \in \[0,1\]$ that captures a "degree of asymmetry": \
   $$
   b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
   $$ {#eq-mixing}
   This means the model now uses three different sets of parameters to describe the spread from the tumor to the LNLs: $b^\text{i}_v$ for the spread to the ipsilateral LNLs, $b_v^{\text{c},\epsilon=\texttt{False}}$ for the spread to the contralateral LNLs as long as the tumor is clearly lateralized, and finally $b_v^{\text{c},\epsilon=\texttt{True}}$ when it crosses the midline. Note, however, that these three sets of spread rates only account for $2 \cdot 2^V + 1$ parameters, since they are coupled via the mixing parameter $\alpha$.

@@

In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters for the full midline model. Considering the parameters of the bilateral model that describe the ipsilateral spread ($b^i_v$, $t_{vr}$ and $p_{adv}$, we note that the bilateral model mostly reproduces the parameter values reported in the earlier publication [] for the unilateral model. The small differences may be mostly due to the differences in training dataset. Therefore, the analysis of the models capability to adequately describe ipsilateral lymph node involvement is not repeated here and we focus instead on the analysis of contralateral involvement. We can further note that the contralateral spread parameters $b^c_v$ are small compared to the ipsilateral parameters $b^i_v$, reflecting the low prevalence of contralateral lymph node involvement for lateralized tumors.

@@

params_table = pd.DataFrame({"Parameter": names, "Mean": means, "Std. Dev.": stds})
(
  params_table.style
  .format("{:.2%}", subset=["Mean"])
  .format("± {:.2%}", subset=["Std. Dev."])
  .apply(right_align, subset=["Mean", "Std. Dev."])
  .hide()
)
```
## Illustration of the model
In this subsection we illustrate parts of the mathematical framework defined in the previous section. In @fig-model-midext-evo (top panel), we visualize the prior distribution over diagnose times $P(t)$. By parameter choice, early T-category tumors are diagnosed on average after 3 time steps. Advanced T-category tumors are diagnosed on average after 4.5 time steps ($p_{adv}=0.45)$, a parameter learned from the data to describe the higher involvement for advanced T-category tumors. The probability per time step for the tumor to grow over the midline, $p_\epsilon = 0.0816$ yields the probability of midline extension $P(\epsilon \mid t)$ at a given time step $t$ (red line). The parameter is adjusted such that the model yields the correct proportion of patients with tumors extending over the midline seen in the dataset. The bottom panel in @fig-model-midext-evo shows the joint probability $P(\epsilon, t)$ to be diagnosed at time step $t$ with a given state of midline extension and T-category.

@@

The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right, \epsilon)$, which is illustrated in @fig-model-state-dist. To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2 \times 2^3 \times 2^3 = 128$ distinct states. LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in table xxx. @fig-model-state-dist shows the joint distribution as two separate heatmaps for the two midline extension states for early/advanced??? T-category tumors. The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%. The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement. The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.

@@

### Contralateral LNL II

@@

### Contralalteral LNL III

@@

### Contralateral LNL IV

>>>>

1. We assume that both ipsilateral and contralateral spread is described through the same graph shown in {#fig-small-graph}.
2. We assume the spread *among* the LNLs to be same on both sides. It is reasonable to assume the lymphatic system is symmetric. Thus, the spread rates from one LNL to the other should be symmetric, too. Formally, this means \
   $$
   \begin{aligned}
   b_v^\text{c} &\neq b_v^\text{i} \\
   t_{rv}^\text{c} &= t_{rv}^\text{i}
   \end{aligned}
   $$ {#eq-symmetries}
   for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
3. The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side. This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to the same position, but on the contralateral side. In the beginning we would have $b_v^\text{c} < b_v^\text{i}$, while in the end, the situation is reversed. If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes. We encode this in a *mixing parameter* $\alpha \in [0,1]$ that captures a "degree of asymmetry": \
   $$
   b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
   $$ {#eq-mixing}
   This means the model now uses three different sets of parameters to describe the spread from the tumor to the LNLs: $b^\text{i}_v$ for the spread to the ipsilateral LNLs, $b_v^{\text{c},\epsilon=\texttt{False}}$ for the spread to the contralateral LNLs as long as the tumor is clearly lateralized, and finally $b_v^{\text{c},\epsilon=\texttt{True}}$ when it crosses the midline. Note, however, that these three sets of spread rates only account for $2 \cdot 2^V + 1$ parameters, since they are coupled via the mixing parameter $\alpha$.

@@

In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters for the full midline model. Considering the parameters of the bilateral model that describe the ipsilateral spread ($b^i_v$, $t_{vr}$ and $p_\text{adv.}$), we note that the bilateral model mostly reproduces the parameter values reported in the earlier publication [@ludwig_modelling_2023] for the unilateral model. The small differences may be mostly due to the differences in training dataset. Therefore, the analysis of the models capability to adequately describe ipsilateral lymph node involvement is not repeated here and we focus instead on the analysis of contralateral involvement. We can further note that the contralateral spread parameters $b^c_v$ are small compared to the ipsilateral parameters $b^i_v$, reflecting the low prevalence of contralateral lymph node involvement for lateralized tumors.

@@

params_table = pd.DataFrame({"Parameter": names, "Mean": means, "Std. Dev.": stds})
(
  params_table.style
  .format("{:.2%}", subset=["Mean"])
  .format("± {:.2%}", subset=["Std. Dev."])
  .apply(right_align, subset=["Mean", "Std. Dev."])
  .hide()
)
```

## Illustration of the model

In this subsection we illustrate parts of the mathematical framework defined in the previous section. In @fig-model-midext-evo (top panel), we visualize the prior distribution over diagnose times $P(t)$. By parameter choice, early T-category tumors are diagnosed on average after 3 time steps. Advanced T-category tumors are diagnosed on average after 4.5 time steps ($p_\text{adv.}=0.45$) <!-- TODO: make dynamic -->, a parameter learned from the data to describe the higher involvement for advanced T-category tumors. The probability per time step for the tumor to grow over the midline, $p_\epsilon = 0.0816$ <!-- TODO: make dynamic --> yields the probability of midline extension $P(\epsilon \mid t)$ at a given time step $t$ (red line). The parameter is adjusted such that the model yields the correct proportion of patients with tumors extending over the midline seen in the dataset. The bottom panel in @fig-model-midext-evo shows the joint probability $P(\epsilon, t)$ to be diagnosed at time step $t$ with a given state of midline extension and T-category.

@@

The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, which is illustrated in @fig-model-state-dist. To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2 \times 2^3 \times 2^3 = 128$ distinct states. LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in table xxx. @fig-model-state-dist shows the joint distribution as two separate heatmaps for the two midline extension states for early/advanced??? T-category tumors. The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%. The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement. The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.

@@

## Contralateral LNL II

@@

## Contralalteral LNL III

@@

## Contralateral LNL IV
